The Categorical Structure of Profunctor Optics

### 2017-08-23 paper.tex
Some cleanup

\todo{TODO: Might have to adjust the entire thing because the action of $\M$ on $\C$ is not strictly monoidal. On the other hand, through the entire Pastro/Street paper pretends associativity/etc is strict, so maybe who cares? What we really want is just a (non-strict) monoidal functor $\M \to [\C, \C]$ from some $\M$.}

@@

\todo{Introduction should include:
\begin{itemize}
\item Previous attempts to generalise optics
\item ...
\end{itemize}
}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be \todo{represented faithfully in $S$},
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

\begin{proposition}
Let $\C$ be a category and $\M$ a monoidal category that acts on $\C$. There is a category $\PreOptic_\M$ that has the same objects as $\C \times \C$, with homsets given by
\begin{align*}
\PreOptic_\M((S, S'), (A, A')) = \int^{M \in \M} \C(S, M A) \times \C(M A', S')
\end{align*}
\end{proposition}
To distinguish these from morphisms in $\C$, we will write preoptics with a crossed arrow: $p : (S, S') \hto (A, A')$.

If $f : S \to M A$ and $g : M A' \to S'$, we write $\langle l, r \rangle$ for their image in $\PreOptic_\M((S, S'), (A, A'))$. We will refer to $M$ as the \emph{complement} for the representative $\langle l, r \rangle$.
\begin{proof}
\todo{Following the doubles paper, this should be the Kleisli category for the monad on $\Prof$ given by:
\begin{align*}
\Phi(S, S', A, A') = \int^{M \in \M} \C(S, M A) \times \C(M A', S')
\end{align*}
}

@@

In \cite{Doubles}, a very similar category was considered. The \emph{double} $\Double$ of a monoidal category $(\C, \otimes, I)$ is a special case of the above, where $\M$ is chosen to be all functors of the form $X \otimes (-)$ for $X \in \C$.

@@

\begin{definition}
An \emph{optic} from $S$ to $A$ is a preoptic $p$ from $(S, S)$ to $(A, A)$ such that $outside(p) = \id_S$, and $inside(p) = \langle \id_{MA} \rangle$ for some $M \in \M$.
\end{definition}

\begin{proposition}
\label{prop-onthenose}
If $p : (S,S) \hto (A,A)$ is an optic, then there exists a representative $p = \langle l, r \rangle$ such that $r \circ l = \id_{MA}$ on the nose for some $M$.
\end{proposition}
\begin{proof}
\todo{This is long and painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

In fact we will prove the more general statement that for $p : (S,S) \hto (A,A)$, if $outside(p) = \id_S$ and $inside(p) = \langle t \rangle$ where $t : MA \to MA$ is idempotent, then we can find a representative $\langle l, r \rangle$ with $l \circ r = t$ on the nose.

@@

Similarly unwinding the coequaliser for $\int^{M \in \M} \C(S, M A) \times \C(M A, S)$, two pairs $(l, r) : (S \to M A, MA \to S)$ and $(l', r') : (S \to NA, NA \to S)$ in $\coprod_{M \in \M} \C(S, M A) \times \C(M A, S)$ are identified when there exists a $\phi : M \to N$ such that
\[
\begin{tikzcd}
M A \ar[r, "\phi_A"] \ar[d, "r", swap] & N A \ar[d, "r'"] \\
S \ar[r, equal] \ar[d, "l", swap] & S \ar[d, "l'"]  \\
MA \ar[r, "\phi_A"] & N A
\end{tikzcd}
\]
commutes. \todo{split the diagram in the middle?}

@@

The above argument has a similar form to those that appear in \cite{OnTheTrace}, which considered (among other things) coends of the form $\int^{c \in \C} \C(c, Fc)$ for an endofunctor $F : \C \to \C$.

In view of this Proposition, whenever we posit a representative $\langle l, r \rangle$ of an optic, we will assume that $l$ and $r$ are mutual two-sided inverses.

\begin{proposition}
There is an subcategory $\Optic_\M$ of $\PreOptic_\M$ given by objects of the form $(X, X)$ and optics between them.
\end{proposition}
\begin{proof}
The identity preoptic $\langle \id_X, \id_X \rangle$ is clearly an optic. Suppose we have two optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$ with complements $M$ and $N$ respectively. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is certainly also an optic:
\begin{align*}
(Ml')l r (Mr') &= (Ml')(Mr') = M(l'r') = \id_{MNA}
\intertext{and}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\end{align*}
\end{proof}

@@

Lenses form the ur-example of optics.

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} in $\C$ is the category of optics with respect to this action: $\Lens = \Optic_\times$.
\end{definition}

@@

Explicitly, given a prelens $\langle l, r \rangle : S \hto A$ these are given by $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Conversely, given $\fget$ and $\fput$, the corresponding element in the coend has representative $\langle l, r \rangle$, where $l = [\id_s, \fget]$ and $r = \fput$. Note that even if the $\fget$ and $\fput$ correspond to a law-abiding lens, this particular representative will not be an isomorphism.

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
If $\langle l, r \rangle : S \hto A$ is a lens then the associated $\fget$ and $\fput$ functions satisfy the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws.
\end{proposition}
\begin{proof}
Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (\pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}

@@

In favourable conditions, we can show the converse holds by constructing a complement. This is similar to the result proved in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

\begin{lemma}
\todo{maybe delete this lemma, not so important}
Suppose $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the $\fget\fput$, $\fput\fget$ and $\fput\fput$ laws. If we have $x_1, x_2 : 1 \to A$. Then the pullbacks $\fget^*(x_1)$ and $\fget^*(x_2)$ are isomorphic.
\end{lemma}
\begin{proof}
We construct a map $f : \fget^*(x_2) \to \fget^*(x_1)$ as the induced map in the diagram
\[
\begin{tikzcd}
\fget^*(x_2) \ar[ddr, bend right = 20] \ar[dr, dashed] \ar[r, "{[x_2^*, x_1 !]}"] & S \times A \ar[dr, "\fput"] & \\
& \fget^*(x_1) \ar[r, swap] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x_1", swap] & A
\end{tikzcd}
\]
That the outside composite commutes is exactly the $\fput\fget$ law. Similarly, we have a map $g : \fget^*(x_1) \to \fget^*(x_2)$.

\todo{they are inverses by the other laws}
\end{proof}

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a lens.
\end{proposition}
\begin{proof}
Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. Now the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M((S, S), (A, A))$. It remains to show that the maps on the left are mutual inverses:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}
Therefore $\langle \fput (i \times A), [j,\fget] \rangle$ is a lens, so  $\langle \fput, [\id_S, \fget] \rangle$ is a lens as we were hoping to show. \todo{maybe do the above in pictures}
\end{proof}

@@

The laws for prisms are:
\begin{align*}
\fmatching \; \freview = \inr \\
[\id_S, \freview] \fmatching = \id_S
\end{align*}
\todo{Are we missing a law?? Analogous to PutPut? I believe we are}
\todo{
\begin{proposition}
When $\C = \Set$, the third law is implied by the other two.
\end{proposition}
}

@@

\todo{
\section{Weaker Condition??}
}
\begin{proposition}
Suppose the pre-lens $\langle l, r \rangle : S \hto A$ satisfies $rl = \id_S$ and $\langle lr \rangle = \langle t \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
This is nearly identical to Proposition \ref{prop-OpticImpliesLensLaws}. As was mentioned in the proof of Proposition \ref{prop-onthenose}, we may assume that $lr = t \times A$ on the nose.

Now, again letting $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$,
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}
\end{proof}

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ obeys the above weaker condition.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
proves that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract?}
\end{proof}

@@

\begin{proposition}
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \PreOptic_\M((S, S), (A, A))
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\PreOptic_\M((S, S), (A, A))
\end{align*}
\todo{This works for $\PreOptic_\M((S, S'), (A, A'))$ just fine, fixing the definition of $E_A$.}
\end{proof}

@@

To show $p$ is monoidal, we must show that both diagrams
\[
\begin{tikzcd}
UP(A,A) \times UQ(A,A) \ar[r, "p_P \times p_Q"] \ar[d, swap] & UP(S, S) \times UQ(S, S)  \ar[d] \\
U(P \odot Q)(A,A)  \ar[r, "p_{P \odot Q}", swap] & U(P \odot Q)(S,S)
\end{tikzcd}
\]
and
\[
\begin{tikzcd}
1 \ar[d] \ar[dr] & \\
UI(A,A)  \ar[r, "p_I", swap] & UI(S,S)
\end{tikzcd}
\]
commute. For the first we have
\begin{align*}
\langle (p_P \times p_Q)(x, y) \rangle &= \langle (UP)(l,r) \alpha_{A,A,M} x,  (UQ)(l,r) \beta_{A,A,M} y\rangle \\
&= \langle (UP)(l,lr) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(by the coend relations \todo{more carefully})} \\
&= \langle (UP)(l,\id_{MA}) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(as $lr = \id_{MA}$)} \\
&= (U(P \odot Q))(l,r) \langle \alpha_{A,A,M} x, \beta_{A,A,M} y \rangle && \text{(by definition of $P \odot Q$)} \\
&= (U(P \odot Q))(l,r) (\alpha_{A,A,M} \odot \beta_{A,A,M}) \langle x, y \rangle && \text{(by definition of $\alpha \odot \beta$)} \\
&= p_{P \odot Q} \langle x, y \rangle
\end{align*}
And for the second:
\begin{align*}
p_I(\id_A) = (UI)(l,r) (M \id_A) = r \id_{MA} l = rl = \id_S
\end{align*}
as required.
\end{proof}

@@

\section{Degenerate Optics}
\todo{The Haskell lens library has some other `optics' that aren't proper optics, but still integrate nicely with the rest of the ecosystem, e.g. Getters, Reviews and Folds}

>>>>

\todo{TODO: Might have to adjust the entire thing because the action of $\M$ on $\C$ is not strictly monoidal. On the other hand, through the entire Pastro/Street paper pretends associativity/etc is strict, so maybe who cares? What we really want is just a (non-strict) monoidal functor $\M \to [\C, \C]$ from some $\M$. I may also want to put $\otimes$ and $\cdot$ everywhere instead of just writing $MA$ and $NMA$.}

@@

\todo{Introduction should include:
\begin{itemize}
\item Previous attempts to generalise optics
\item Some remark about how it's hard to isolate exactly what $\Set$-like properties are being used
\end{itemize}
}
A lens from $S$ to $A$ is a pair of maps $\fget : S \to A$ and $\fput : S \times A \to S$ that typically obeys three laws (whose names are in diagrammatic order):
\begin{itemize}
\item ``$\fput\fget$'': $\fget \; \fput = \pi_2$ as maps $S \times A \to A$, so that any update to $A$ should be \todo{represented faithfully in $S$},
\item ``$\fget\fput$'': $\fput [\id_S, \fget] = \id_S$ as maps $S \to S$, so that if $A$ is not changed then $S$ should not change; and,
\item ``$\fput\fput$'': $\fput (\fput \times \id_A) = \fput \, \pi_{1, 3}$ as maps $S \times A \times A \to S$, so any update to $A$ completely overwrites previous updates.
\end{itemize}

@@

\begin{proposition}
Let $\C$ be a category and $(\M, \otimes, I)$ a monoidal category that acts on $\C$. There is a category $\PreOptic_\M$ that has the same objects as $\C \times \C$, with homsets given by
\begin{align*}
\PreOptic_\M((S, S'), (A, A')) = \int^{M \in \M} \C(S, M A) \times \C(M A', S')
\end{align*}
\end{proposition}
To distinguish these from morphisms in $\C \times \C$, we will write preoptics with a crossed arrow: $p : (S, S') \hto (A, A')$.

If $f : S \to M A$ and $g : M A' \to S'$, we write $\langle l, r \rangle$ for their image in $\PreOptic_\M((S, S'), (A, A'))$.
\begin{proof}
\todo{Following the doubles paper, this should be the Kleisli category for the monad on $\Prof$ given by:
\begin{align*}
\Phi(S, S', A, A') = \int^{M \in \M} \C(S, M A) \times \C(M A', S')
\end{align*}
}

@@

In \cite{Doubles}, a very similar category was considered. The \emph{double} $\Double$ of a monoidal category $(\C, \otimes, I)$ is a special case of the above, where $\C$ acts on itself by $\otimes$.

@@

\begin{definition}
An \emph{optic} from $S$ to $A$ is a preoptic $p$ from $(S, S)$ to $(A, A)$ such that $outside(p) = \id_S$, and $inside(p) = \langle t A \rangle$ for some $t : M \to M \in \M$.
\end{definition}

\begin{proposition}
\label{prop-onthenose}
If $p : (S,S) \hto (A,A)$ is an optic, then there exists a representative $p = \langle l, r \rangle$ such that $rl = \phi A$ on the nose for some $\phi : M \to M$.
\end{proposition}
\begin{proof}
\todo{This is painful but seems important. There may be a more general principle at work here but I couldn't see it.}

@@

@@

Similarly unwinding the coequaliser for $\int^{M \in \M} \C(S, M A) \times \C(M A, S)$, two pairs $(l, r) : (S \to M A, MA \to S)$ and $(l', r') : (S \to NA, NA \to S)$ in $\coprod_{M \in \M} \C(S, M A) \times \C(M A, S)$ are identified when there exists a $\phi : M \to N$ such that
\[
\begin{tikzcd}
M A \ar[r, "\phi_A"] \ar[d, "r", swap] & N A \ar[d, "r'"] \\
S \ar[r, equal] \ar[d, "l", swap] & S \ar[d, "l'"]  \\
MA \ar[r, "\phi_A"] & N A
\end{tikzcd}
\]
commutes. \todo{put the $S$ on the outside?}

@@

The above argument has a similar form to some that appear in \cite{OnTheTrace}, which considered (among other things) coends of the form $\int^{c \in \C} \C(c, Fc)$ for an endofunctor $F : \C \to \C$.

In view of this Proposition, whenever we posit a representative $\langle l, r \rangle$ of an optic, we may assume that $lr = \phi A$ on the nose, for some $\phi : M \to M$.

\begin{proposition}
There is an subcategory $\Optic_\M$ of $\PreOptic_\M$ given by objects of the form $(X, X)$ and optics between them.
\end{proposition}
\begin{proof}
The identity preoptic $\langle \id_X : X \to IX, \id_X : IX \to X \rangle$ is clearly an optic.

Suppose we have two optics $\langle l, r \rangle : R \hto S$ and $\langle l', r' \rangle : S \hto A$, where $l : R \to MS$ and $l' : S \to NA$, and such that $lr = \phi S$ and $l'r' = \psi A$. Then the composite $\langle (Ml')l, r (Mr')  \rangle$ is also an optic:
\begin{align*}
r (Mr')(Ml')l &= r M(r'l') l = rl = \id_S
\intertext{and}
(Ml')l r (Mr') &= (Ml')(\phi S)(Mr') = (\phi NA) M(l')M(r') = (\phi NA) M(l'r') = (\phi NA) M(\psi A) = \phi(\psi A)
\end{align*}
by the naturality of the action of $\M$.
\end{proof}

@@

Lenses are the ur-example of a class of optics.

\begin{definition}
Suppose $\C$ has finite products, and let $\C$ act on itself via the bifunctor $\times : \C \times \C \to \C$. The \emph{category of lenses} $\Lens$ in $\C$ is the category of optics with respect to this action: $\Lens = \Optic_\times$.
\end{definition}

@@

Explicitly this isomorphism states that, given a prelens $\langle l, r \rangle : S \hto A$, the corresponding $\fget$ and $\fput$ functions are given by $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. In the other direction, given $\fget$ and $\fput$, the corresponding element in the coend has representative $\langle l, r \rangle$, where $l = [\id_s, \fget]$ and $r = \fput$. Note that even if the $\fget$ and $\fput$ correspond to a law-abiding lens, this particular representative will not be an isomorphism.

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
Suppose $\langle l, r \rangle : S \hto A$ is a pre-lens such that $rl = \id_S$ and $\langle lr \rangle = \langle \phi \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$, for some $\phi : M \to M$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
By Proposition \ref{prop-onthenose}, we may assume that $lr = \phi \times A$ on the nose.

Let $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$. Verification of the laws is straightforward:
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}

@@

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ is a prelens.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract?}
\end{proof}

In favourable conditions, we can upgrade this result to show that a lens $S \hto A$ in fact implies that $S \cong M \times A$ for some $M$. This is similar to the result proved in \cite[Corollary 13]{AlgebrasAndUpdateStrategies} for $\Set$.

\begin{proposition}
Suppose $\C$ has finite limits and that there is a morphism $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then there exists $C \in \C$ with $S \cong C \times A$.
\end{proposition}
\begin{proof}
\todo{TODO: This isn't really what we want to show, we want to start from an optic, not from something obeying the $\fget/\fput$ laws.}

Set $C$ to be the pullback of $\fget$ along $x$, so there is a map $i : C \to S$ with $\fget \, i = x !_C$. There is also a map $j : S \to C$ induced by the following diagram:
\[
\begin{tikzcd}
S \ar[ddr, bend right = 20] \ar[dr, "j", dashed] \ar[r, "{[\id_S, x !_S]}"] & S \times A \ar[dr, "\fput"] & \\
& C \ar[r, "i"] \ar[d] \arrow[dr, phantom, "\lrcorner", very near start] & S \ar[d, "\fget"] \\
& 1\ar[r, "x", swap] & A
\end{tikzcd}
\]
which commutes by the $\fput\fget$ law. Note that $ji = \id_C$ by the universal property of pullbacks. The claim is that $\fput (i \times A) : C \times A \to S$ and $[j,\fget] : S \to C \times A$ are mutual inverses. This is easily checked:
\begin{align*}
\fput (i \times A)[j,\fget] &= \fput [ij,\fget] \\
&= \fput [\fput [\id_S, x!_S],\fget] && \text{(by definition of $j$)} \\
&= \fput [\id_S,\fget] && \text{(by $\fput\fput$)} \\
&= \id_S && \text{(by $\fget\fput$)}
\intertext{and}
[j,\fget]\fput (i \times A) &= [j\fput (i \times A),\fget\,\fput (i \times A)] && \text{(by universal property of product)} \\
&= [j\fput (i \times A), \pi_2 (i \times A)] && \text{(by $\fput\fget$)} \\
&= [j\fput (i \times A), \pi_2] && \\
&= [jij\fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S,x !_S] \fput (i \times A), \pi_2] && \\
&= [j\fput [\id_S, x !_S] \pi_1 (i \times A), \pi_2] && \text{(by $\fput\fput$ \todo{todo: detail})}\\
&= [jij \pi_1 (i \times A), \pi_2] && \\
&= [jiji \pi_1, \pi_2] && \\
&= [\pi_1, \pi_2] && \\
&= \id_{C \times A}
\end{align*}

Additionally, the commutative diagram
\[
\begin{tikzcd}
C \times A \ar[r, "i \times A"] \ar[d, "\fput (i \times A)", swap] & S \times A \ar[d, "\fput"] \\
S \ar[r, equal] \ar[d, "{[j,\fget]}", swap] & S \ar[d, "{[\id_S, \fget]}"]  \\
C \times A \ar[r, "i \times A", swap] & S \times A
\end{tikzcd}
\]
is a witness that $\langle \fput (i \times A), [j,\fget] \rangle = \langle \fput, [\id_S, \fget] \rangle$ as elements of $\PreOptic_\M((S, S), (A, A))$.
\end{proof}

@@

The laws for prisms are the obvious duals to the lens laws. 
\begin{align*}
\fmatching \; \freview &= \inr \\
[\id_S, \freview] \fmatching &= \id_S \\
(\fmatching \sqcup \id_A) \fmatching &= \mathrm{in}_{1,3} \, \fmatching
\end{align*}
\todo{
\begin{proposition}
When $\C = \Set$, and probably more generally, the third law is implied by the other two.
\end{proposition}
}

@@

@@

\begin{proposition}
\label{prop-profunctor-optics-are-optics}
\begin{align*}
[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) &\cong \PreOptic_\M((S, S), (A, A))
\end{align*}
\end{proposition}
\begin{proof}
We have the chain of isomorphisms:
\begin{align*}
&[\Tamb_\C(\M), \Set]((U-)(A,A),(U-)(S,S)) \\
\cong \;&[\Tamb_\C(\M), \Set](\Tamb_\C(\M)(\Pastro_\M E_A, -), (U-)(S,S))  && \text{(by Lemma \ref{lemma-rep})}\\
\cong \;&(U\Pastro_\M E_A)(S,S)  && \text{(by Yoneda)} \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times E_A(C,D) \times \C(MD,S) \\
= \;&\int^{M \in \M} \int^{C,D \in \C} \C(S, MC) \times \C(C,A) \times \C(A,D) \times \C(MD,S) \\
\cong \;&\int^{M \in \M} \C(S, MA) \times \C(MA, S)  && \text{(by co-Yoneda twice)} \\
= \;&\PreOptic_\M((S, S), (A, A))
\end{align*}
\todo{This works for $\PreOptic_\M((S, S'), (A, A'))$ just fine, just define $E_{A,A'}$ in the obvious way.}
\end{proof}

@@

To show $p$ is monoidal, we must show that both diagrams
\[
\begin{tikzcd}
UP(A,A) \times UQ(A,A) \ar[r, "p_P \times p_Q"] \ar[d, swap] & UP(S, S) \times UQ(S, S)  \ar[d] \\
U(P \odot Q)(A,A)  \ar[r, "p_{P \odot Q}", swap] & U(P \odot Q)(S,S)
\end{tikzcd}
\]
and
\[
\begin{tikzcd}
1 \ar[d] \ar[dr] & \\
UI(A,A)  \ar[r, "p_I", swap] & UI(S,S)
\end{tikzcd}
\]
commute. For the first we have
\begin{align*}
\langle (p_P \times p_Q)(x, y) \rangle &= \langle (UP)(l,r) \alpha_{A,A,M} x,  (UQ)(l,r) \beta_{A,A,M} y\rangle \\
&= \langle (UP)(l,lr) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(by the coend relations)} \\
&= \langle (UP)(l,\phi A) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(as $lr = \phi A$)} \\
&= \langle (UP)((\phi A)l,\id_{MA}) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(by the coend relations again)} \\
&= \langle (UP)(lrl,\id_{MA}) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(as $lr = \phi A$ again)} \\
&= \langle (UP)(l,\id_{MA}) \alpha_{A,A,M} x, (UQ)(\id_{MA},r) \beta_{A,A,M} y\rangle && \text{(as $rl = \id_S$)} \\
&= (U(P \odot Q))(l,r) \langle \alpha_{A,A,M} x, \beta_{A,A,M} y \rangle && \text{(by definition of $P \odot Q$)} \\
&= (U(P \odot Q))(l,r) (\alpha_{A,A,M} \odot \beta_{A,A,M}) \langle x, y \rangle && \text{(by definition of $\alpha \odot \beta$)} \\
&= p_{P \odot Q} \langle x, y \rangle
\end{align*}
And for the second:
\begin{align*}
p_I(\id_A) = (UI)(l,r) (M \id_A) = r \id_{MA} l = rl = \id_S
\end{align*}
as required.
\end{proof}

@@

\todo{
\section{Degenerate Optics}
The Haskell lens library has some other `optics' that aren't proper optics, but still integrate nicely with the rest of the ecosystem, e.g. Getters, Reviews and Folds. I am starting to doubt those fit into this framework though.}
